---
title: "Comprehensive Analysis of EMS Arrival Time Influence on Fatality Severity"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)  
library(lubridate)  
library(ggplot2)  
library(broom)    
library(gridExtra)    
```

## 1. Introduction
This analysis addresses the question:
- How does the EMS arrival time influence the severity of fatal outcomes?

We will explore how EMS response times correlate with the severity of fatal accidents by categorizing accidents into different severity levels based on the number of fatalities. This analysis will involve descriptive statistics, visual analysis, and advanced statistical testing.

## 2. Data Loading and Preparation

### 2.1 Data Loading
We start by loading the dataset and examining the structure and summary of the data.

```{r data-loading}
# Load the dataset
fars_data <- read.csv("NTAD_Fatality_Analysis_Reporting_System_2022_Accidents_3270373163975898856.csv")

# Overview of the dataset
glimpse(fars_data)
summary(fars_data)
```

### 2.2 Data Preparation
In this step, we will prepare the data:
- **EMS Response Time Calculation**: Calculate the total EMS arrival time in minutes using `ARR_HOUR` and `ARR_MIN`.
- **Conversion of Fatality Count**: Ensure the `FATALS` column is numeric for analysis.
- **Categorize Fatality Severity**: Create a new column for severity levels based on the number of fatalities.

```{r data-preparation}
# Calculate EMS arrival time in minutes
fars_data$EMS_ARRIVAL_TIME <- (as.numeric(fars_data$ARR_HOUR) * 60) + as.numeric(fars_data$ARR_MIN)

# Convert FATALS column to numeric
fars_data$FATALS <- as.numeric(fars_data$FATALS)

# Categorize fatal accidents into severity levels
fars_data$SEVERITY_LEVEL <- cut(
  fars_data$FATALS,
  breaks = c(0, 1, 3, 5, Inf),
  labels = c("Low (1 Fatality)", "Medium (2-3 Fatalities)", "High (4-5 Fatalities)", "Very High (6+ Fatalities)")
)

# Overview of the EMS-related data after preparation
glimpse(fars_data[, c("EMS_ARRIVAL_TIME", "FATALS", "SEVERITY_LEVEL")])
summary(fars_data[, c("EMS_ARRIVAL_TIME", "FATALS", "SEVERITY_LEVEL")])
```

## 3. Exploratory Data Analysis (EDA)

### 3.1 Descriptive Statistics
We will examine basic statistics for EMS arrival times and fatalities, including measures of central tendency and variability.

```{r descriptive-statistics}
# Summary statistics for EMS arrival time and fatalities
ems_stats <- fars_data %>%
  summarize(
    mean_arrival_time = mean(EMS_ARRIVAL_TIME, na.rm = TRUE),
    median_arrival_time = median(EMS_ARRIVAL_TIME, na.rm = TRUE),
    sd_arrival_time = sd(EMS_ARRIVAL_TIME, na.rm = TRUE),
    min_arrival_time = min(EMS_ARRIVAL_TIME, na.rm = TRUE),
    max_arrival_time = max(EMS_ARRIVAL_TIME, na.rm = TRUE),
    mean_fatalities = mean(FATALS, na.rm = TRUE),
    max_fatalities = max(FATALS, na.rm = TRUE)
  )
print(ems_stats)
```

### 3.2 EMS Arrival Time by Severity Level
We will compare EMS response times for different severity levels to see if delays correlate with more severe outcomes.

```{r severity-analysis}
# Boxplot of EMS arrival time by severity level
ggplot(fars_data, aes(x=SEVERITY_LEVEL, y=EMS_ARRIVAL_TIME, fill=SEVERITY_LEVEL)) +
  geom_boxplot(alpha=0.6) +
  labs(title="EMS Arrival Time by Fatality Severity Level",
       x="Severity Level",
       y="EMS Arrival Time (minutes)",
       fill="Severity Level") +
  theme_minimal()
```

### 3.3 Correlation Analysis of Severity
We will explore the correlation between EMS arrival times and the severity of accidents.

```{r severity-correlation}
# Correlation analysis between EMS arrival time and number of fatalities
correlation_severity <- cor.test(fars_data$EMS_ARRIVAL_TIME, fars_data$FATALS, use="complete.obs")
cat("Correlation between EMS Arrival Time and Severity (Number of Fatalities):\n")
print(correlation_severity)
```

### 3.4 Linear Regression Analysis
Perform a linear regression to determine if EMS arrival time is a predictor of the number of fatalities.

```{r linear-regression}
# Linear regression model: Predicting fatalities based on EMS arrival time
linear_model <- lm(FATALS ~ EMS_ARRIVAL_TIME, data = fars_data)
summary(linear_model)

# Visualizing the regression line
ggplot(fars_data, aes(x = EMS_ARRIVAL_TIME, y = FATALS)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Linear Regression: EMS Arrival Time vs Number of Fatalities",
       x = "EMS Arrival Time (minutes)",
       y = "Number of Fatalities") +
  theme_minimal()
```

## 4. Advanced Statistical Analysis

### 4.1 ANOVA for Severity Levels
We will use an ANOVA test to see if there's a statistically significant difference in EMS arrival times across different severity levels.

```{r anova-analysis}
# ANOVA to compare means of EMS arrival times across severity levels
anova_result <- aov(EMS_ARRIVAL_TIME ~ SEVERITY_LEVEL, data = fars_data)
summary(anova_result)

# Post-hoc test to determine which groups differ
TukeyHSD(anova_result)
```

### 4.2 Logistic Regression Analysis
We will perform logistic regression to see if EMS arrival time influences the probability of a high-severity outcome.

```{r logistic-regression}
# Creating a binary variable for high severity
fars_data$HIGH_SEVERITY <- ifelse(fars_data$FATALS > 3, 1, 0)

# Logistic regression model
logistic_model <- glm(HIGH_SEVERITY ~ EMS_ARRIVAL_TIME, data = fars_data, family = binomial)
summary(logistic_model)

# Exponentiating coefficients to interpret them as odds ratios
exp(coef(logistic_model))
```

## 5. Insights and Recommendations

### Summary of Findings
Based on the analysis, summarize the key observations and their implications for EMS response effectiveness.

```{r summary-findings}
cat("Summary of Findings:\n")
if (correlation_severity$p.value < 0.05) {
  cat("- The analysis indicates a significant relationship between EMS arrival time and the severity of accidents.\n")
  if (correlation_severity$estimate > 0) {
    cat("- Longer EMS response times are associated with more severe fatal outcomes.\n")
  } else {
    cat("- Shorter EMS response times are associated with less severe fatal outcomes.\n")
  }
} else {
  cat("- There is no significant relationship between EMS arrival time and the severity of accidents.\n")
}
```

