---
title: "Dead_Ends"
author: "Team5"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
runtime: shiny
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# ***Dead Ends and Data Trends: A Journey Through America's Recent Accident Stats***

### An insightful journey into understanding the patterns behind road fatalities in the U.S. and creating actionable insights for a safer future.


# ***Project Overview***

### Motor vehicle accidents are a leading cause of unintentional injury-related deaths in the U.S.  Using the 2022 FARS dataset, our analysis focuses on revealing trends and risk factors that contribute to fatal crashes.


### Our first step is getting ready by loading the necessary packages and the data. 


```{r, include=T, results='asis',message=TRUE}
# Loading the necessary libraries
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(plotly)
```


```{r, include=T, results='markup',message=TRUE}
# Loading the dataset
dataset <- read_csv("dataset.csv")

# Now viewing the structure of the dataset
str(dataset)
summary(dataset)
```


## Next, we check for duplicate rows. Let's clear them out!
```{r, include=T, results='markup',message=TRUE}
dataset <- dataset %>% distinct()
```


## Moving on to handling missing values...
```{r, include=T, results='markup',message=TRUE}
missing_values <- colSums(is.na(dataset))
print(missing_values[missing_values > 0])
```



```{r, include=T, results='markup',message=TRUE}
# Lets rename columns for consistency
colnames(dataset)<- str_to_lower(colnames(dataset))
colnames(dataset) <- str_replace_all(colnames(dataset), " ", "_")
colnames(dataset)
print(ncol(dataset))
```



```{r, include=T, results='markup',message=TRUE}
road_data <- dataset %>% 
  select(rur_urbname) %>%
  drop_na() %>%
  filter(!rur_urbname %in% c("Unknown", "Not Reported", "Trafficway Not in State Inventory")) %>%
  mutate(
    rur_urbname = factor(rur_urbname, labels = c("Rural", "Urban")),
  )
```

## Let's summarize the data to calculate the total fatalities by state.

```{r, include=T, results='markup',message=TRUE}
state_county_summary <- dataset %>%
  group_by(statename, countyname) %>%
  summarise(Total_Fatalities = sum(fatals)) %>%
  ungroup()

state_summary <- state_county_summary %>%
  group_by(statename) %>%
  summarise(Total_Fatalities = sum(Total_Fatalities)) %>%
  arrange(desc(Total_Fatalities))

```

```{r, include=T, results='markup',message=TRUE}

library(shiny)
ui <- fluidPage(
  titlePanel("Drill-Down Analysis of Crash Fatalities by State and County"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput("state_select", "Select a State:", choices = unique(state_summary$statename), selected = "Texas")
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel("State-Level Analysis", plotlyOutput("statePlot")),  # State-level plot tab
        tabPanel("County-Level Analysis", plotlyOutput("countyPlot")) # County-level plot tab
      )
    )
  )
)

# Define server logic for the Shiny app
server <- function(input, output) {
  
  # State-level Plot for Top 10 States by Total Fatalities
  output$statePlot <- renderPlotly({
    top_state_summary <- state_summary %>%
      slice_head(n = 10)  # Select top 10 states by total fatalities
    
    state_bar_plot <- ggplot(top_state_summary, aes(x = reorder(statename, Total_Fatalities), y = Total_Fatalities, fill = Total_Fatalities)) +
      geom_bar(stat = "identity", color = "black", width = 0.7) +
      geom_text(aes(label = Total_Fatalities), hjust = -0.2, color = "black", size = 4) +
      labs(title = "Top 10 States by Total Fatalities",
           x = "State",
           y = "Total Number of Fatalities") +
      theme_minimal() +
      scale_fill_gradient(low = "lightblue", high = "darkred") +
      theme(
        plot.title = element_text(hjust = 0.5, size = 16),
        legend.position = "none"
      ) +
      coord_flip()  # Horizontal bar plot
    
    ggplotly(state_bar_plot, tooltip = c("x", "y"))
  })
  
  # County-level Plot for Selected State
  output$countyPlot <- renderPlotly({
    # Filter county-level data for the selected state
    county_data <- state_county_summary %>%
      filter(statename == input$state_select) %>%
      arrange(desc(Total_Fatalities)) %>%
      slice_head(n = 10)  # Select top 10 counties by fatalities within the selected state
    
    county_bar_plot <- ggplot(county_data, aes(x = reorder(countyname, Total_Fatalities), y = Total_Fatalities, fill = Total_Fatalities)) +
      geom_bar(stat = "identity", color = "black", width = 0.7) +
      geom_text(aes(label = Total_Fatalities), hjust = -0.2, color = "black", size = 4) +
      labs(title = paste("Top 10 Counties by Total Fatalities in", input$state_select),
           x = "County",
           y = "Total Number of Fatalities") +
      theme_minimal() +
      scale_fill_gradient(low = "lightyellow", high = "darkorange") +
      theme(
        plot.title = element_text(hjust = 0.5, size = 16),
        legend.position = "none"
      ) +
      coord_flip()  # Horizontal bar plot for counties
    
    ggplotly(county_bar_plot, tooltip = c("x", "y"))
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)

```





## **3. What is the relationship between weather and lighting conditions on the occurrence of fatal accidents?**

## 3.1 **Bivariate Analysis**

```{r, include=T, results='markup',message=TRUE}

# Define UI for the Shiny app
ui <- fluidPage(
  titlePanel("Fatalities by Weather and Lighting Conditions"),
  
  sidebarLayout(
    sidebarPanel(
      helpText("Explore fatalities by weather and lighting conditions"),
      hr()
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel("Weather Conditions", plotlyOutput("weatherPlot")),
        tabPanel("Lighting Conditions", plotlyOutput("lightingPlot"))
      )
    )
  )
)

# Define server logic for Shiny app
server <- function(input, output) {
  
  # Filter and aggregate data for weather conditions
  weather_data <- reactive({
    dataset %>%
      filter(weathername != "Not Reported" & weathername != "Reported as Unknown") %>%
      group_by(weathername) %>%
      summarise(Total_Fatalities = sum(fatals, na.rm = TRUE)) %>%
      arrange(desc(Total_Fatalities))
  })
  
  # Filter and aggregate data for lighting conditions
  lighting_data <- reactive({
    dataset %>%
      filter(lgt_condname != "Not Reported" & lgt_condname != "Reported as Unknown") %>%
      group_by(lgt_condname) %>%
      summarise(Total_Fatalities = sum(fatals, na.rm = TRUE)) %>%
      arrange(desc(Total_Fatalities))
  })
  
  # Plot for Weather Conditions with Log Scale and Original Value Labels
  output$weatherPlot <- renderPlotly({
    weather_plot <- ggplot(weather_data(), aes(x = reorder(weathername, Total_Fatalities), y = Total_Fatalities)) +
      geom_bar(stat = "identity", fill = "skyblue") +
      coord_flip() +
      scale_y_log10() +  # Apply logarithmic scale to y-axis
      labs(title = "Fatalities by Weather Conditions", x = "Weather Condition", y = "Total Fatalities (Log Scale)") +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5)) +
      geom_text(aes(label = scales::comma(Total_Fatalities)),  # Display original values on bars
                hjust = -0.2, size = 3, color = "black")  # Adjust positioning for readability
    
    ggplotly(weather_plot, tooltip = c("x", "y"))
  })
  
  # Plot for Lighting Conditions with Log Scale and Original Value Labels
  output$lightingPlot <- renderPlotly({
    lighting_plot <- ggplot(lighting_data(), aes(x = reorder(lgt_condname, Total_Fatalities), y = Total_Fatalities)) +
      geom_bar(stat = "identity", fill = "orange") +
      coord_flip() +
      scale_y_log10() +  # Apply logarithmic scale to y-axis
      labs(title = "Fatalities by Lighting Conditions", x = "Lighting Condition", y = "Total Fatalities (Log Scale)") +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5)) +
      geom_text(aes(label = scales::comma(Total_Fatalities)),  # Display original values on bars
                hjust = -0.2, size = 3, color = "black")  # Adjust positioning for readability
    
    ggplotly(lighting_plot, tooltip = c("x", "y"))
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)

## Filtering out the records that is unknown or not reported
dataset <- dataset %>%
  filter(!(weathername %in% c("Not Reported", "Reported as Unknown")))

dataset <- dataset %>%
  filter(!(lgt_condname %in% c("Not Reported", "Reported as Unknown")))
```

#### Clear weather has the highest number of fatalities, likely due to high exposure as most driving happens in clear conditions.

#### Cloudy and Rain also pose significant risks, while severe conditions like Snow, Fog, and Crosswinds still contribute to fatalities, albeit at lower rates.

#### This insight helps emphasize that while severe weather conditions contribute to road fatalities, most incidents occur during clear weather, likely due to the high volume of travel under these conditions.

## 3.2 **Multivariate analysis** : Let's analyze fatality counts by both weather and lighting conditions using a heatmap

```{r, include=T, results='markup',message=TRUE}
# Aggregating fatal counts by weather and lighting
weather_light <- dataset %>%
  group_by(weathername, lgt_condname) %>%
  summarize(total_fatals = sum(fatals, na.rm = TRUE))

# Interactive Heatmap
heatmap_plot <- plot_ly(weather_light, x = ~weathername, y = ~lgt_condname, z = ~total_fatals, type = "heatmap") %>%
  layout(title = "Fatality Count by Weather and Lighting Condition",
         xaxis = list(title = "Weather Condition"),
         yaxis = list(title = "Lighting Condition"))

heatmap_plot
```

#### **Clear and Daylight Conditions**: These conditions have the highest fatality counts, likely due to the higher volume of traffic under these conditions.

#### **Clear and Dark - Not Lighted**: The relatively high fatality count here suggests the need for improved street lighting on roads frequently used at night.

#### **Weather Impact**: Rain and Cloudy conditions contribute to fatalities across various lighting conditions, pointing to the increased risk associated with low-visibility and wet surfaces.


### ***This heatmap can help guide targeted safety measures:***

#### **Enhanced Lighting**: Roads that are dark and not lighted could benefit from additional lighting to reduce nighttime crashes, especially under clear conditions.

#### **Weather-Specific Warnings**: Increased signage or public warnings during rainy or cloudy weather may help drivers exercise caution, particularly in low-light conditions.


## **3.3 Location-Based Analysis**

### **Top 10 States by Fatal Accidents in Adverse Weather or Poor Lighting**
```{r, include=T, results='markup',message=TRUE}
state_analysis <- dataset %>%
  filter(weathername %in% c("Rain", "Sleet or Hail", "Blowing Snow","Blowing Sand,Soil Dirt","Severe Crosswinds","Fog,Smog,Smoke","Freezing Rain or Drizzle","Snow"), lgt_condname %in% c("Dark-Not Lighted", "Dark - Lighted","Dark- Unknown Lighting")) %>%
  group_by(statename) %>%
  summarise(Total_Fatalities = sum(fatals, na.rm = TRUE)) %>%
  arrange(desc(Total_Fatalities))


# Select the top 10 states with the highest fatalities
top_adverse_states <- head(state_analysis, 10)  # Top 10 states

ggplot(top_adverse_states, aes(x = reorder(statename, Total_Fatalities), y = Total_Fatalities, fill = Total_Fatalities)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Top 10 States by Fatal Accidents in Adverse Weather or Poor Lighting",
       x = "State", y = "Total Fatalities") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10),  # Increase font size for readability
        legend.position = "none")  # Hide the legend
```

#### This chart shows the top 10 U.S. states with the highest number of fatal accidents occurring under adverse weather or poor lighting conditions. The states are ordered by the total number of fatalities, with Florida and Texas experiencing the highest counts. 

#### **Future Implications:** This analysis helps identify locations that could benefit from targeted safety measures in adverse weather or poor lighting, such as improved lighting infrastructure, better signage, or increased public awareness campaigns during severe weather events.

### **Top 10 States by Fatal Accidents in Clear Weather and Good lighting**

```{r, include=T, results='markup',message=TRUE}
state_analysis <- dataset %>%
  filter(weathername %in% c("Clear","Cloudy"), lgt_condname %in% c("Daylight", "Dawn","Dusk")) %>%
  group_by(statename) %>%
  summarise(Total_Fatalities = sum(fatals, na.rm = TRUE)) %>%
  arrange(desc(Total_Fatalities))


# Select the top 10 states with the highest fatalities
top_states <- head(state_analysis, 10)  # Top 10 states

ggplot(top_states, aes(x = reorder(statename, Total_Fatalities), y = Total_Fatalities, fill = Total_Fatalities)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "States with highest Fatal Accidents in clear weather and good lighting",
       x = "State", y = "Total Fatalities") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10),  # Increase font size for readability
        legend.position = "none")  # Hide the legend
```

#### Despite favorable weather and lighting conditions, fatalities remain high in clear weather in states like California, Texas, and Florida. This suggests that driver behavior, traffic congestion, and infrastructure quality could be key factors influencing accident rates in these states.

#### However, Focused interventions, such as better road lighting, clear weather warnings, and infrastructure improvements to handle severe weather, could benefit states like Florida, Texas, and New York.


## **3.4 Statistical Tests**
#### Let us support the above results by performing statistical tests:

#### **ANOVA (Analysis of Variance) test** followed by a Tukey's HSD  post hoc test to analyze the impact of lighting conditions (lgt_condname) on the fatal accident counts (fatals) 

```{r, include=T, results='markup',message=TRUE}

anova_lighting <- aov(fatals ~ lgt_condname, data = dataset)

# Display ANOVA summary
summary(anova_lighting)
```

#### Let us define our null and alternate hypothesis

#### **Null Hypothesis (H0)**: There is no difference in the mean number of fatalities across different lighting conditions (i.e., lighting conditions do not affect fatal accident counts).

#### **Alternative Hypothesis (H1)**: There is a difference in the mean number of fatalities across at least one pair of lighting conditions.


#### ***p-value:***

#### The p-value for lgt_condname is 0.0054, which is below the common significance level of 0.05.

#### This indicates that we reject the null hypothesis and conclude that there is a statistically significant difference in the mean fatal accident counts across different lighting conditions.

#### ***F-Statistic:***

#### The F-value of 2.72 suggests that the variance between the groups (lighting conditions) is more than what would be expected by chance, reinforcing that lighting conditions do affect fatal accident counts.

#### ***Sum of Squares:***

#### The sum of squares (SS) values indicate the variance within the groups (Residuals) and the variance attributed to the lighting conditions (lgt_condname).

#### The ANOVA results indicate that lighting conditions significantly impact the number of fatal accidents, meaning that certain lighting conditions are associated with different fatality rates.

#### Next Step with **Tukey's HSD Test**: Since the ANOVA indicates a significant effect, the Tukey's HSD test can be used to pinpoint which specific lighting conditions differ from each other in terms of fatal accident counts. 

```{r, include=T, results='markup',message=TRUE}
tukey_results <- TukeyHSD(anova_lighting, "lgt_condname")
print(tukey_results)

```

### ***Significant Comparisons:***

#### Only the comparison "Dark - Not Lighted" vs. "Dark - Lighted" has a p-value of 0.001, which is below 0.05. This suggests a statistically significant difference in mean fatalities between "Dark - Not Lighted" and "Dark - Lighted" conditions.

#### The positive diff value (0.02067) indicates that "Dark - Not Lighted" has a higher mean fatality count than "Dark - Lighted".


### ***Practical Implications :***

#### "Dark - Not Lighted" vs. "Dark - Lighted": Since there is a statistically significant difference between these two conditions, with "Dark - Not Lighted" associated with higher fatality counts, it may suggest that insufficient lighting in dark conditions contributes to a higher risk of fatal accidents.

#### **Recommendation**: Improving lighting in areas that are "Dark - Not Lighted" may help reduce fatal accidents, as lighting seems to play a crucial role in accident prevention under dark conditions.

#### For other lighting conditions, the lack of significant differences suggests that fatalities do not vary considerably across these conditions. This could imply that lighting in conditions like Dawn, Daylight, and Dusk might be sufficient, and improvements in these areas may not have as strong an impact on reducing fatal accidents.


#### **Independent t-test** (assuming the fatality rates in adverse and clear weather conditions are from two separate groups).

```{r, include=T, results='markup',message=TRUE}

# Filter data for adverse and clear weather conditions
adverse_conditions <- c("Rain", "Sleet or Hail", "Blowing Snow","Blowing Sand,Soil Dirt","Severe Crosswinds","Fog,Smog,Smoke","Freezing Rain or Drizzle","Snow")
fair_conditions <- c("Clear","Cloudy")

# Summarize fatalities by state for adverse weather conditions
adverse_weather <- dataset %>%
  filter(weathername %in% adverse_conditions) %>%
  group_by(statename) %>%
  summarise(Adverse_Fatalities = sum(fatals, na.rm = TRUE))

# Summarize fatalities by state for fair weather condition
fair_weather <- dataset %>%
  filter(weathername == fair_conditions) %>%
  group_by(statename) %>%
  summarise(Clear_Fatalities = sum(fatals, na.rm = TRUE))

# Merge the data for adverse and clear weather fatalities by state
weather_comparison <- merge(adverse_weather, fair_weather, by = "statename", all = TRUE)

# Replace NA values with 0 (in case a state has no fatalities in one of the conditions)
weather_comparison[is.na(weather_comparison)] <- 0

# Perform the independent t-test
t_test_result <- t.test(weather_comparison$Adverse_Fatalities, weather_comparison$Clear_Fatalities, 
                        alternative = "two.sided", var.equal = FALSE)

print(t_test_result)
```

#### **p-value**:

#### The p-value is 1e-05, which is extremely low (below the standard significance level of 0.05).
#### This indicates that we can reject the null hypothesis and conclude that there is a statistically significant difference in fatality rates between adverse weather and clear weather conditions.

#### **t-Statistic**:

#### The t-value of -5 suggests a substantial difference between the two groups. The negative value indicates that the mean fatality rate in adverse weather is lower than that in clear weather.

#### The results hence indicate a significant difference in fatality rates between adverse and clear weather conditions, with higher fatalities observed in clear weather.

#### This outcome could suggest that during clear weather, there may be higher traffic volumes, faster speeds, or other risk factors that increase the likelihood of fatalities, whereas adverse weather might lead drivers to exercise more caution, resulting in fewer fatalities.

#### Hence it is important to underastand that, in addition to weather and lighting conditions, there are several other critical factors like road infrastructure, juction types and driver behaviour which can equally influence the likelihood and severity of accidents. Addressing these factors can provide a more holistic approach to improving road safety and reducing fatalities.



