---
title: "Analysis of Trends in Fatal Accidents"
Team: "5"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)  
library(lubridate) 
library(ggplot2)  
```

## 1. Introduction (SMART Question 4)
This analysis addresses one critical SMART question:
4. Are there noticeable trends in fatal accidents throughout the year 2022, and has the number of fatalities been increasing or decreasing in certain regions?

We will be using the 2022 Fatality Analysis Reporting System (FARS) dataset to explore this question, focusing on trends.

## 2. Data Loading and Preparation

```{r data-loading}
# Load the dataset
fars_data <- read.csv("NTAD_Fatality_Analysis_Reporting_System_2022_Accidents_3270373163975898856.csv")

glimpse(fars_data)
summary(fars_data)
```


### Dataset Overview
```{r}
cat("Dataset Overview:\n")
cat("- Number of records:", nrow(fars_data), "\n")
cat("- Number of variables:", ncol(fars_data), "\n")
cat("- Key Variables:\n")
print(names(fars_data))
```


### Data Preparation
- **Year Extraction**: If the dataset contains a 'Date' or 'Year' column, ensure proper formatting.
- **Filtering Relevant Columns**: Focus on columns relevant to year, region, and fatality status.

```{r data-preparation}
if ("DATE" %in% names(fars_data)) {
  fars_data$DATE <- as.Date(fars_data$DATE, format="%Y-%m-%d")
  fars_data$MONTH <- month(fars_data$DATE, label = TRUE)
}

glimpse(fars_data)
summary(fars_data)
```

## Graphical Representations

```{r graphical-representations}
fars_data$total_fatalities <- ifelse(fars_data$FATALS > 0, 1, 0)

# Histogram of Total Fatalities
ggplot(fars_data, aes(x=total_fatalities)) +
    geom_histogram(binwidth=1, fill="blue", color="black", alpha=0.7) +
    labs(title="Histogram of Total Fatalities", x="Total Fatalities", y="Frequency") +
    theme_minimal()

# Boxplot of Fatalities by Month
ggplot(fars_data, aes(x=MONTH, y=total_fatalities)) +
    geom_boxplot(fill="lightgreen") +
    labs(title="Boxplot of Fatalities by Month", x="Month", y="Total Fatalities") +
    theme_minimal()
```


## Normality Tests

```{r normality-tests}
# Check the structure of the dataset
str(fars_data)

# Create total_fatalities based on the FATALS column
fars_data$total_fatalities <- fars_data$FATALS

# Check the number of non-missing values
non_missing_count <- sum(!is.na(fars_data$total_fatalities))
cat("Number of non-missing values in total_fatalities:", non_missing_count, "\\n")

# Display unique values and summary of total_fatalities
unique_values <- unique(fars_data$total_fatalities)
summary(fars_data$total_fatalities)
cat("Unique values in total_fatalities:", unique_values, "\\n")

# Run Shapiro-Wilk Normality Test
if (non_missing_count >= 3 && non_missing_count <= 5000) {
    shapiro_test <- shapiro.test(fars_data$total_fatalities)
    cat("Shapiro-Wilk Normality Test:\\n")
    print(shapiro_test)
} else {
    cat("Insufficient sample size for Shapiro-Wilk test.\\n")
}

# Q-Q Plot for Normality Check
qqnorm(fars_data$total_fatalities, main="Q-Q Plot for Total Fatalities")
qqline(fars_data$total_fatalities, col="red")
```

## Statistical Test

### Chi-Square Test
```{r chi-square-test}
# Chi-Square Test
table_states <- table(fars_data$STATENAME)
chi_square_test <- chisq.test(table_states)
cat("Chi-Square Test Results:\n")
print(chi_square_test)
```


## 3. Monthly trends for 2022


# Analyzing fatalities by month to identify seasonal patterns
```{r}
monthly_summary <- fars_data %>% 
  group_by(MONTH) %>% 
  summarize(total_fatalities = n(), .groups = 'drop')

ggplot(monthly_summary, aes(x=MONTH, y=total_fatalities)) +
  geom_line(group=1, color="blue", size=1.2) +
  geom_point(color="red", size=3) +
  scale_x_continuous(breaks=1:12) +  
  labs(title="Trends in Fatal Accidents by Month (2022)",
       x="Month",
       y="Total Fatalities") +
  theme_minimal()
```

### Regional Analysis for the year 2022

```{r, fig.width=14, fig.height=10}
state_monthly_summary <- fars_data %>% 
  group_by(STATENAME, MONTH) %>% 
  summarize(total_fatalities = n(), .groups = 'drop')

print(head(state_monthly_summary))

ggplot(state_monthly_summary, aes(x=MONTH, y=total_fatalities, color=STATENAME, group=STATENAME)) +
  geom_line(size=1, alpha=0.7) +  
  geom_point(size=2, alpha=0.8) +  
  scale_x_continuous(breaks=1:12) + 
  scale_y_continuous(breaks = seq(0, max(state_monthly_summary$total_fatalities, na.rm = TRUE), by = 10)) + 
  labs(title="State-wise Trends in Fatal Accidents by Month",
       x="Month",
       y="Total Fatalities",
       color="State") +
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.grid.minor = element_blank()) 


ggsave("enlarged_state_trends.png", width = 18, height = 12, dpi = 300)
```
## 4. Insights and Recommendation


# Summarize the monthly and regional trends
```{r}
peak_month <- monthly_summary$MONTH[which.max(monthly_summary$total_fatalities)]

cat("Summary of Findings:\n")
cat("- The month with the highest number of fatal accidents in 2022 was", peak_month, ".\n")

```

# Identify the region with the highest fatalities in each month

```{r}
regional_monthly_summary <- fars_data %>% 
  group_by(STATENAME, MONTH) %>% 
  summarize(total_fatalities = n(), .groups = 'drop')

high_risk_states <- regional_monthly_summary %>% 
  group_by(MONTH) %>% 
  slice(which.max(total_fatalities)) %>% 
  select(MONTH, STATENAME, total_fatalities)

cat("States with the highest number of fatalities for each month:\n")
print(high_risk_states)
```


#Conclusion for the following question

#Checking if the no.of fatalities have been decreasing or increasing over the months in different regions

```{r}
regional_trends <- fars_data %>% 
  group_by(STATENAME, MONTH) %>% 
  summarize(total_fatalities = n(), .groups = 'drop')

trend_direction <- regional_trends %>%
  group_by(STATENAME) %>%
  summarize(trend = cor(MONTH, total_fatalities, use="complete.obs")) %>%
  mutate(direction = ifelse(trend > 0, "Increasing", "Decreasing"))

cat("Trend Analysis Summary for Fatalities in 2022:\n")
print(trend_direction)
```
