---
title: "Analysis of Trends in Fatal Accidents"
Team: "5"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)  
library(lubridate) 
library(ggplot2)  
```

## 1. Introduction (SMART Question 4)
This analysis addresses one critical SMART question:
4. Are there noticeable trends in fatal accidents throughout the year 2022, and has the number of fatalities been increasing or decreasing in certain regions?

We will be using the 2022 Fatality Analysis Reporting System (FARS) dataset to explore this question, focusing on trends.

## 2. Data Loading and Preparation

```{r data-loading}
# Load the dataset
fars_data <- read_csv("/Users/jahnaviv/Documents/GitHub/DATS_6101_TEAM5/dataset/cleaned_FARS.csv")

glimpse(fars_data)
summary(fars_data)
```


### Dataset Overview
```{r}
cat("Dataset Overview:\n")
cat("- Number of records:", nrow(fars_data), "\n")
cat("- Number of variables:", ncol(fars_data), "\n")
cat("- Key Variables:\n")
print(names(fars_data))
```


### Data Preparation
- **Year Extraction**: If the dataset contains a 'Date' or 'Year' column, ensure proper formatting.
- **Filtering Relevant Columns**: Focus on columns relevant to year, region, and fatality status.

```{r data-preparation}
if ("DATE" %in% names(fars_data)) {
  fars_data$DATE <- as.Date(fars_data$DATE, format="%Y-%m-%d")
  fars_data$MONTH <- month(fars_data$DATE, label = TRUE)
}

glimpse(fars_data)
summary(fars_data)
```

## Graphical Representations

```{r graphical-representations}
# Create total_fatalities based on the fatals column
fars_data$total_fatalities <- ifelse(fars_data$fatals > 0, 1, 0)

# Histogram of Total Fatalities
ggplot(fars_data, aes(x=total_fatalities)) +
    geom_histogram(binwidth=1, fill="blue", color="black", alpha=0.7) +
    labs(title="Histogram of Total Fatalities", x="Total Fatalities (0 = No, 1 = Yes)", y="Frequency") +
    theme_minimal()

# Boxplot of Fatalities by Month
ggplot(fars_data, aes(x=factor(monthname, levels=month.name), y=fatals)) +  # Convert monthname to a factor
    geom_boxplot(fill="lightgreen") +
    labs(title="Boxplot of Fatalities by Month", x="Month", y="Total Fatalities") +
    theme_minimal()
```

## Normality Tests

```{r normality-tests}
str(fars_data)

fars_data$total_fatalities <- fars_data$fatals 

non_missing_count <- sum(!is.na(fars_data$total_fatalities))
cat("Number of non-missing values in total_fatalities:", non_missing_count, "\\n")

unique_values <- unique(fars_data$total_fatalities)
summary(fars_data$total_fatalities)
cat("Unique values in total_fatalities:", unique_values, "\\n")

if (non_missing_count >= 3 && non_missing_count <= 5000) {
    shapiro_test <- shapiro.test(fars_data$total_fatalities)
    cat("Shapiro-Wilk Normality Test:\\n")
    print(shapiro_test)
} else {
    cat("Insufficient sample size for Shapiro-Wilk test.\\n")
}

qqnorm(fars_data$total_fatalities, main="Q-Q Plot for Total Fatalities")
qqline(fars_data$total_fatalities, col="red")
```

## Statistical Test

### Chi-Square Test
```{r chi-square-test}
table_states <- table(fars_data$statename)

chi_square_test <- chisq.test(table_states)

cat("Chi-Square Test Results:\n")
print(chi_square_test)
```


## 3. Monthly trends for 2022


# Analyzing fatalities by month to identify seasonal patterns
```{r}
monthly_summary <- fars_data %>% 
  group_by(monthname) %>% 
  summarize(total_fatalities = sum(fatals, na.rm = TRUE), .groups = 'drop') 

monthly_summary$monthname <- factor(monthly_summary$monthname, levels = month.name)

ggplot(monthly_summary, aes(x=monthname, y=total_fatalities)) +
  geom_line(group=1, color="blue", size=1.2) +
  geom_point(color="red", size=3) +
  labs(title="Trends in Fatal Accidents by Month (2022)",
       x="Month",
       y="Total Fatalities") +
  theme_minimal()
```

### Regional Analysis for the year 2022

```{r, fig.width=14, fig.height=10}
# Summarizing total fatalities by state and month
state_monthly_summary <- fars_data %>% 
  group_by(statename, monthname) %>%  # Ensure correct column names
  summarize(total_fatalities = sum(fatals, na.rm = TRUE), .groups = 'drop')

# Display the head of the summary
print(head(state_monthly_summary))

# Create a factor for monthname to ensure it is ordered correctly
state_monthly_summary$monthname <- factor(state_monthly_summary$monthname, levels = month.name)

# Plotting state-wise trends in fatalities by month
ggplot(state_monthly_summary, aes(x=monthname, y=total_fatalities, color=statename, group=statename)) +
  geom_line(size=1, alpha=0.7) +  
  geom_point(size=2, alpha=0.8) +  
  scale_y_continuous(breaks = seq(0, max(state_monthly_summary$total_fatalities, na.rm = TRUE), by = 10)) + 
  labs(title="State-wise Trends in Fatal Accidents by Month",
       x="Month",
       y="Total Fatalities",
       color="State") +
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.grid.minor = element_blank()) 

# Save the plot as a PNG file
ggsave("enlarged_state_trends.png", width = 18, height = 12, dpi = 300)
```

# Identify the region with the highest fatalities in each month

```{r}
regional_monthly_summary <- fars_data %>% 
  group_by(statename, monthname) %>%  
  summarize(total_fatalities = sum(fatals, na.rm = TRUE), .groups = 'drop')

high_risk_states <- regional_monthly_summary %>% 
  group_by(monthname) %>% 
  slice(which.max(total_fatalities)) %>% 
  select(monthname, statename, total_fatalities)

cat("States with the highest number of fatalities for each month:\n")
print(high_risk_states)
```


#Conclusion for the following question

#Checking if the no.of fatalities have been decreasing or increasing over the months in different regions

```{r}
regional_trends <- fars_data %>% 
  group_by(statename, monthname) %>%
  summarize(total_fatalities = sum(fatals, na.rm = TRUE), .groups = 'drop')

regional_trends <- regional_trends %>%
  mutate(month_numeric = match(monthname, month.name)) 

trend_direction <- regional_trends %>%
  group_by(statename) %>%
  filter(n() > 1) %>% 
  summarize(trend = cor(month_numeric, total_fatalities, use = "complete.obs")) %>%
  mutate(direction = ifelse(trend > 0, "Increasing", "Decreasing"))

cat("Trend Analysis Summary for Fatalities in 2022:\n")
print(trend_direction)
```
