---
title: "Dead_Ends"
author: "Team5"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
runtime: shiny
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# ***Dead Ends and Data Trends: A Journey Through America's Recent Accident Stats***

### An insightful journey into understanding the patterns behind road fatalities in the U.S. and creating actionable insights for a safer future.


# ***Project Overview***

### Motor vehicle accidents are a leading cause of unintentional injury-related deaths in the U.S.  Using the 2022 FARS dataset, our analysis focuses on revealing trends and risk factors that contribute to fatal crashes.


### Our first step is getting ready by loading the necessary packages and the data. 


```{r, include=T, results='asis',message=TRUE}
# Loading the necessary libraries
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(plotly)
```


```{r, include=T, results='markup',message=TRUE}
# Loading the dataset
dataset <- read_csv("dataset.csv")

# Now viewing the structure of the dataset
str(dataset)
summary(dataset)
```


## Next, we check for duplicate rows. Let's clear them out!
```{r, include=T, results='markup',message=TRUE}
dataset <- dataset %>% distinct()
```


## Moving on to handling missing values...
```{r, include=T, results='markup',message=TRUE}
missing_values <- colSums(is.na(dataset))
print(missing_values[missing_values > 0])
```


## There are way too many NA values in "TWAY_ID2", "X" AND "Y". Let's drop those columns. *We are dropping a lot many other columns as they are not needed.*

```{r, include=T, results='markup',message=TRUE}
dataset <- dataset %>% select(-x, -y, -STATE, -CITY, -COUNTY, -MONTH, -DAYNAME, -DAY_WEEK, -MINUTENAME, -TWAY_ID2, -ROUTE, -RUR_URB, -FUNC_SYS, -RD_OWNER, -NHS, -SP_JUR, -LATITUDENAME, -LONGITUDNAME, -MILEPT, -HARM_EV, -MAN_COLL, -MILEPTNAME, -RELJCT1, -RELJCT2, -TYP_INT, -REL_ROAD, -WRK_ZONE, -LGT_COND, -WEATHER, -SCH_BUS, -RAIL, -NOT_MIN, -ARR_MINNAME, -ARR_HOUR, -HOSP_MN, -SCH_BUSNAME, -RAILNAME, -PERNOTMVIT, -VE_FORMS, -PERSONS)

# Showing the remaining features.

# Lets rename columns for consistency
colnames(dataset)<- str_to_lower(colnames(dataset))
colnames(dataset) <- str_replace_all(colnames(dataset), " ", "_")
colnames(dataset)
```


## Saving the new and cleaned dataset...

#```{r, include=T, results='markup',message=TRUE}
#write.csv(dataset, "cleaned_dataset.csv", row.names = FALSE)
#```

```{r, include=T, results='markup',message=TRUE}
road_data <- dataset %>% 
  select(rur_urbname) %>%
  drop_na() %>%
  filter(!rur_urbname %in% c("Unknown", "Not Reported", "Trafficway Not in State Inventory")) %>%
  mutate(
    rur_urbname = factor(rur_urbname, labels = c("Rural", "Urban")),
  )
```

## Let's summarize the data to calculate the total fatalities by state.

```{r, include=T, results='markup',message=TRUE}
state_county_summary <- dataset %>%
  group_by(statename, countyname) %>%
  summarise(Total_Fatalities = sum(fatals)) %>%
  ungroup()

state_summary <- state_county_summary %>%
  group_by(statename) %>%
  summarise(Total_Fatalities = sum(Total_Fatalities)) %>%
  arrange(desc(Total_Fatalities))

```

```{r, include=T, results='markup',message=TRUE}

library(shiny)
ui <- fluidPage(
  titlePanel("Drill-Down Analysis of Crash Fatalities by State and County"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput("state_select", "Select a State:", choices = unique(state_summary$statename), selected = "Texas")
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel("State-Level Analysis", plotlyOutput("statePlot")),  # State-level plot tab
        tabPanel("County-Level Analysis", plotlyOutput("countyPlot")) # County-level plot tab
      )
    )
  )
)

# Define server logic for the Shiny app
server <- function(input, output) {
  
  # State-level Plot for Top 10 States by Total Fatalities
  output$statePlot <- renderPlotly({
    top_state_summary <- state_summary %>%
      slice_head(n = 10)  # Select top 10 states by total fatalities
    
    state_bar_plot <- ggplot(top_state_summary, aes(x = reorder(statename, Total_Fatalities), y = Total_Fatalities, fill = Total_Fatalities)) +
      geom_bar(stat = "identity", color = "black", width = 0.7) +
      geom_text(aes(label = Total_Fatalities), hjust = -0.2, color = "black", size = 4) +
      labs(title = "Top 10 States by Total Fatalities",
           x = "State",
           y = "Total Number of Fatalities") +
      theme_minimal() +
      scale_fill_gradient(low = "lightblue", high = "darkred") +
      theme(
        plot.title = element_text(hjust = 0.5, size = 16),
        legend.position = "none"
      ) +
      coord_flip()  # Horizontal bar plot
    
    ggplotly(state_bar_plot, tooltip = c("x", "y"))
  })
  
  # County-level Plot for Selected State
  output$countyPlot <- renderPlotly({
    # Filter county-level data for the selected state
    county_data <- state_county_summary %>%
      filter(statename == input$state_select) %>%
      arrange(desc(Total_Fatalities)) %>%
      slice_head(n = 10)  # Select top 10 counties by fatalities within the selected state
    
    county_bar_plot <- ggplot(county_data, aes(x = reorder(countyname, Total_Fatalities), y = Total_Fatalities, fill = Total_Fatalities)) +
      geom_bar(stat = "identity", color = "black", width = 0.7) +
      geom_text(aes(label = Total_Fatalities), hjust = -0.2, color = "black", size = 4) +
      labs(title = paste("Top 10 Counties by Total Fatalities in", input$state_select),
           x = "County",
           y = "Total Number of Fatalities") +
      theme_minimal() +
      scale_fill_gradient(low = "lightyellow", high = "darkorange") +
      theme(
        plot.title = element_text(hjust = 0.5, size = 16),
        legend.position = "none"
      ) +
      coord_flip()  # Horizontal bar plot for counties
    
    ggplotly(county_bar_plot, tooltip = c("x", "y"))
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)

```

## 4. **How do road conditions (urban/rural, road type, and traffic controls) impact the severity of crashes?** 

### Before analyzing this, let's prep the data

```{r, include=T, results='markup',message=TRUE}

road_fatalities <- dataset %>%
  filter(func_sysname != "Unknown" & func_sysname != "Not Reported" & func_sysname != "Trafficway Not in State Inventory") %>%
  select(func_sysname, fatals) %>%
  drop_na() %>%
  mutate(func_sysname = as.factor(func_sysname))

fatality_summary <- road_fatalities %>%
  group_by(func_sysname) %>%
  summarise(
    Total_Fatalities = sum(fatals),
    SD_Fatalities = sd(fatals),  # Standard deviation for reference
    Count = n()
  )

# Print summary for inspection
print(fatality_summary)

```

## 4.1 Now analyzing the total fatalities by the road type...

```{r, include=T, results='markup',message=TRUE}

# Sort the summary data by Total Fatalities for better visual clarity
fatality_summary <- fatality_summary %>%
  arrange(desc(Total_Fatalities))

# Enhanced bar plot with additional features
fatality_total_bar_plot <- ggplot(fatality_summary, aes(x = reorder(func_sysname, -Total_Fatalities), y = Total_Fatalities, fill = Total_Fatalities)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  geom_text(aes(label = Total_Fatalities), vjust = -0.5, color = "black", size = 3.5) +  # Annotations for exact values
  labs(title = "Total Fatalities by Road Type",
       x = "Road Type",
       y = "Total Number of Fatalities") +
  theme_minimal() +
  scale_fill_gradient(low = "lightblue", high = "darkred") +  # Color gradient for emphasis
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

fatality_total_bar_plotly <- ggplotly(fatality_total_bar_plot, tooltip = c("x", "y", "Total_Fatalities")) %>%
  layout(
    xaxis = list(title = "Road Type"),
    yaxis = list(title = "Total Number of Fatalities"),
    width = 900,  # Adjust width for readability
    height = 500  # Adjust height for a balanced look
  )

fatality_total_bar_plotly
```


### 4.2 **Since we're now dealing with total counts rather than averages, a chi-square test can be more suitable than an ANOVA to determine if there are statistically significant differences in the distribution of fatalities across road types.**


```{r, include=T, results='markup',message=TRUE}

fatality_table <- table(road_fatalities$func_sysname, road_fatalities$fatals > 0)

chi_square_result <- chisq.test(fatality_table)
print(chi_square_result)

```

## Understanding the results...

#### A large chi-square value, like 36993, indicates a large deviation from this "no-difference" scenario, suggesting that some road types are clearly associated with more fatalities than others.

#### p-value < 2e-16 - A p-value less than 0.05 is typically considered statistically significant, meaning the observed differences are unlikely to be due to random chance. Here, the p-value is much smaller (effectively zero), meaning the differences we're seeing are almost certainly real and not due to random chance.

## What are some of the practical implications?

#### Practical Implications

#### **High-Risk Road Types:**

#### Since fatalities are not randomly distributed, certain road types are inherently riskier. This could mean that these roads need extra safety measures.For example, if highways and major arterials show higher fatalities, these road types might benefit from stricter speed controls, increased patrolling, and improved road signage.

#### **Data-Driven Intervention:**

#### The statistical significance (p-value < 2e-16) gives us confidence that investing in safety improvements on high-fatality road types is likely to make a measurable impact on reducing fatalities. This helps decision-makers focus resources where they are most needed rather than spreading resources equally across all road types.

#### **Targeted Policies:**

#### With evidence that fatalities cluster on certain road types, policies could be customized. For example, highways might benefit from median barriers and crash cushions, while local roads might benefit from better lighting and pedestrian crossings.


## 4.3 **Now, let's identify the top fatal routes by routename and functional system name.**

```{r, include=T, results='markup',message=TRUE}

route_function_summary <- dataset %>%
  group_by(routename, func_sysname) %>%
  summarise(fatals = n()) %>%
  arrange(desc(fatals)) %>%
  ungroup()

# Filter the top 10 high-frequency routes and functional systems
top_route_function_summary <- route_function_summary %>%
  slice_head(n = 10)

```

```{r, include=T, results='markup',message=TRUE}
crash_freq_plot <- ggplot(top_route_function_summary, aes(x = reorder(routename, fatals), y = fatals, fill = func_sysname)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  labs(title = "Top Routes and Functional Systems by Crash Frequency",
       x = "Route Name",
       y = "Fatals") +
  theme_minimal() +
  scale_fill_brewer(palette = "Paired") +  # Colorful palette for road types
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    legend.title = element_text(size = 10),
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  coord_flip()  # Flip for better readability


crash_freq_plotly <- ggplotly(crash_freq_plot, tooltip = c("x", "y", "fill"))

crash_freq_plotly
```

```{r, include=T, results='markup',message=TRUE}
crash_freq_heatmap <- ggplot(top_route_function_summary, aes(x = func_sysname, y = reorder(routename, fatals), fill = fatals)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkred", name = "Crash Count") +
  labs(title = "Crash Frequency Heatmap by Route and Functional System",
       x = "Functional System",
       y = "Route Name") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )

crash_freq_heatmap_plotly <- ggplotly(crash_freq_heatmap, tooltip = c("x", "y", "fill"))

crash_freq_heatmap_plotly
```

## **Insights**

#### **High Traffic Volume Routes are High-Risk**: Routes like State Highway, US Highway, and Interstate show high crash frequencies, likely due to high traffic volume and higher speeds. This suggests that major highways and state routes may need targeted safety interventions to manage these risks effectively.

#### **Diverse Functional Systems on Certain Routes**: The presence of multiple functional systems (e.g., Principal Arterial and Minor Arterial) on State Highways and Local Streets indicates that these roads experience varied traffic conditions, which could contribute to higher crash risks.

## **Potential for Safety Interventions**:

#### State and US Highways could benefit from interventions such as speed control measures, improved signage, and roadway design enhancements. Local streets and county roads, while having fewer crashes overall, may still need safety measures, especially in urban areas with pedestrian traffic.


## 4.4 **Do fatalities and frequency differ significantly between urban and rural functional systems?**

```{r, include=T, results='markup',message=TRUE}

# Summarize crash severity and frequency by functional system and urban/rural classification
summary_data <- dataset %>%
  filter(rur_urbname != "Unknown" & rur_urbname != "Not Reported" & rur_urbname != "Trafficway Not in State Inventory") %>%
  group_by(rur_urbname, func_sysname) %>%
  summarise(
    Avg_Severity = mean(fatals, na.rm = TRUE),
    Crash_Frequency = n()
  ) %>%
  ungroup()

ui <- fluidPage(
  titlePanel("Comparing Crash Characteristics on Urban vs. Rural Functional Systems"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput("system_select", "Select Functional System:", choices = unique(summary_data$func_sysname)),
      hr(),
      helpText("Select a functional system to compare crash severity and frequency in urban vs. rural areas.")
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel("Crash Severity", plotlyOutput("severityPlot")),
        tabPanel("Crash Frequency", plotlyOutput("frequencyPlot")),
        tabPanel("Statistical Test Results", verbatimTextOutput("testResults"))
      )
    )
  )
)

server <- function(input, output){
  
  # Filter data based on selected functional system
  filtered_data <- reactive({
    summary_data %>%
      filter(func_sysname == input$system_select)
  })
  
  # Plot: Average Crash Severity in Urban vs. Rural Areas
  output$severityPlot <- renderPlotly({
    severity_plot <- ggplot(filtered_data(), aes(x = rur_urbname, y = Avg_Severity, fill = rur_urbname)) +
      geom_bar(stat = "identity", position = "dodge", color = "black") +
      labs(title = paste("Average Crash Severity in", input$system_select),
           x = "Area Type (Urban/Rural)",
           y = "Average Crash Severity") +
      scale_fill_manual(values = c("Urban" = "skyblue", "Rural" = "salmon")) +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5))
    
    ggplotly(severity_plot, tooltip = c("x", "y"))
  })
  
  # Plot: Crash Frequency in Urban vs. Rural Areas
  output$frequencyPlot <- renderPlotly({
    frequency_plot <- ggplot(filtered_data(), aes(x = rur_urbname, y = Crash_Frequency, fill = rur_urbname)) +
      geom_bar(stat = "identity", position = "dodge", color = "black") +
      labs(title = paste("Crash Frequency in", input$system_select),
           x = "Area Type (Urban/Rural)",
           y = "Crash Frequency") +
      scale_fill_manual(values = c("Urban" = "lightgreen", "Rural" = "darkorange")) +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5))
    
    ggplotly(frequency_plot, tooltip = c("x", "y"))
  })
}

shinyApp(ui = ui, server = server)
```
